### 2017-02-16 real/draft.tex
[draft] Polish "numerical implementation"
Remove several comments

\fred{Check below if \refeq{spk} used with flat prior}

@@

We return to the general formulation of \refeq{sll} and re-arrange its
factors into
\begin{align}
  \lleq{asc}
  p(Q\cond n,\bml)
  &= \int \rmdx{\lambda}\; \rmdx{\bmphi}\;
    p(\lambda\cond n)\;
    p(\bmphi\cond n,\bml)\;
    p(Q\cond \lambda,\bmphi), \\
  \lleq{asd}
  p(Q\cond \lambda,\bmphi)
  &= \sum_N
    p(N\cond \lambda)\;
    \int \rmdx{\bmL}\;
    p(Q\cond N,\bmL)\;
    p(\bmL\cond N,\bmphi),
\end{align}
with $p(\bmL\cond N,\bmphi) = \prod_{j=1}^N p(L_j\cond\bmphi)$ as
before. Again all quantities are related to a single bin in frequency.
Writing the moments and variance of $L$ as\footnote{The $k$-th moment
  of variable $x$ is defined as the expectation value $\mu_k =
  \expect{x^k} = \int \rmdx{x}\,x^k\,p(x),\; k = 0,1,2,\ldots$.} %
\begin{align}
  \lleq{ram}
  \mu =  \mu_1 &\equiv \expect{\Lum}, \qquad
  \mu_2 \equiv \expect{\Lum^2}, \\
  \sigma^2 &\equiv \variance{\Lum}
  = \expect{\Lum^2} - \expect{\Lum}^2,
\end{align}
it is easy to show that, for the Poissonian compound process
(\ref{asd}), the first moment and variance of $Q$ are related to those
of $L$ by
%%%%%%%%%%%%%%%%
\fredmargin{citation} %
%%%%%%%%%%%%%%%
\begin{align}
  \lleq{raa}
  \expect{Q} = \lambda \mu, \qquad
  \variance{Q} = \lambda \mu_2 = \lambda (\mu^2 + \sigma^2) \,.
\end{align}
for any probability $p(L\cond\bmphi)$.
%
Higher-order quantities such as the skewness and kurtosis\footnote{
  The variance is the second so-called \textit{cumulant}. The relation
  for all orders is that the $k$-th order cumulant $\kappa_q(Q)$ is
  equal to $\lambda$ times the $k$-th order moment of $L$,
  $\mu_k$. The skewness of $p(Q\cond \lambda,\bmphi)$ is therefore
  $\kappa_3(Q)/\kappa_2(Q)^{3/2} = \lambda \mu_3/(\lambda \mu_2)^{3/2}
  = \lambda^{-1/2} \mu_3/\mu_2^{3/2}$.}  %
are suppressed by powers of $\lambda^{-1/2}$.
%%%%%%%%%%%
\hans{H comment (not for text): The magnitude of the correction to
  Eq.~(\ref{rab}) due to $\kappa_3$ and higher orders is unclear.}
%%%%%%%%%%%
From Theorem 4.3.1 of Ref.~\cite{bening2002generalized}, we conclude
that the asymptotic distribution of $\Lumtot$ when
$\lambda \to \infty$, given the first and second moment of $\Lum$, is
a Gaussian or normal distribution,
\begin{align}
  \lleq{rab}
  p(\Lumtot \cond \lambda,\bmphi)
  &= \GaussianDist \left( \Lumtot \cond \expect{Q}, \variance{Q} \right) %
    = \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\mu^2 {+} \sigma^2) \right).
\end{align}
This drastic simplification eliminates both the sum over $N$ and the
integral over $\bmL$ in \refeq{asd}.

It remains to find the integrals over $\lambda$ and $\bmphi$ in
\refeq{asc}. As before, the posterior for $\lambda$ is the Gamma
distribution of \refeq{pms}. %
%
To determine $p(\bmphi\cond n,\bml)$, we note that the result
(\ref{rab}) also simplifies the calculation of the parameter posterior
$p(\bmphi\cond n,\bml)$ since, independently of the actual functional
model for $p(L\cond\bmphi)$ the only relevant information is contained
in the first moment $\mu$ and variance $\sigma^2$.  As set out in
Jaynes~\cite[Ch. 7.10]{jaynes2003probability}, knowledge of $\mu$ and
$\sigma^2$ implies that we can assign a Gaussian distribution to
$p(L\cond\bmphi)$
\begin{align}
  \lleq{rae}
  p(\Lum \cond \bmphi) =
  p(\Lum \cond \mu, \sigma^2 ) &= \GaussianDist(\Lum \cond \mu, \sigma^2)
\end{align}
even though the actual data histograms may not be Gaussian. The reason
is that a Gaussian represents our state of knowledge \emph{before}
data $\bmy$ is observed. With this assignment,
$\bmphi =(\mu,\sigma^2)$ and the posterior becomes, using Bayes'
Theorem,
\begin{align}
  \lleq{rac}
  p(\mu,\sigma^2\cond n,\bml)
  &= \frac{
    \GaussianDist(\bml\cond\mu,\sigma^2)\;
    p(\mu\cond n)\;
    p(\sigma^2\cond n)}
  {p(\bml\cond n)}
\end{align}
\hans{I have inserted a uniform prior for $\mu$ here; see comments
  below. The original Gaussian prior has been retained below but I
  would suggest that the compliations introduced by it are
  unnecessary.}  %
In keeping with prior information that luminosities are always
positive, we assign an improper uniform prior
$p(\mu\cond\mu_{\rm max}) = 1/\mu_{\rm max}$ independent of $n$; as
always, the posterior is independent of
$\mu_{\rm max}$.
For $\sigma^2$, we use an $n$-independent Inverse Gamma
distribution prior with hyperparameters $(a_0,b_0)$ which reads
\begin{align}
  \lleq{rad}
  p(\sigma^2\cond a_0,b_0)
  = \InvGammaDist(\sigma^2\cond a_0,b_0)
  &= \frac{b_0^{a_0}}{\Gamma(a_0)} (\sigma^2)^{-a_0-1} \exp(-b_0/\sigma^2)
\end{align}
and is ``conjugate'' to the normal distribution likelihood of
\refeq{rae} in that the posterior for $(\mu,\sigma^2)$ is also a
Gaussian in $\mu$ and an Inverse Gamma in $\sigma^2$, but with
parameters updated by the data,
\begin{align}
  \lleq{rbd}
  p(\mu,\sigma^2\cond n,\bml,a_0,b_0)
  &= \GaussianDist(\mu\cond \mu_n,\sigma_n^2)
  \cdot
  \InvGammaDist(\sigma^2\cond a_n,b_n) \\
  \mu_n &= \langle\ell\rangle\\
  \sigma_n^2 &= \sigma^2/n \\
  a_n &= a_0 + \tfrac{1}{2}n \\
  b_n &= b_0 + \tfrac{1}{2} n\left( \langle\ell^2\rangle - \langle\ell\rangle^2\right)
\end{align}
with data sample moments $\langle\lum^k\rangle$ defined as $(1/n)
\sum_{j=1}^n \lum_j^k$.
%
Inserting Eqs.~(\ref{pms}) and (\ref{rbd}) into \refeq{asc}, we obtain
\begin{align}
  \lleq{ral}
  p(\Lumtot \cond n,\bml)
  &= \int \rmdx{\lambda}\; \rmdx{\mu}\; \rmdx{\sigma^2}\;
  \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\sigma^2{+}\mu^2) \right)
  \times\\
  &\quad\times
  \GammaDist(\lambda \cond n{-}a{+}1 , 1) \cdot
  \GaussianDist(\mu\cond \mu_n,\sigma_n^2) \cdot
  \InvGammaDist(\sigma^2\cond a_n,b_n) \;
  \nonumber
\end{align}
While the integral cannot be carried out analytically, the integrand
is unimodal and strongly peaked, so that the Laplace or saddle-point
approximation~\cite[Ch. 27]{mackay2003information} can be applied.

@@

In \reffig{asymptotic}, we contrast the asymptotic expression with the
``exact'' expression \refeq{spk} where the integrals over $\bmphi$ are
done numerically to a requested relative accuracy of $10^{-5}$ through
the Cubature
package\cite{cubature,genz1980remarks,berntsen1991adaptive}. We
computed $P(\Lumtot \cond n, \bml)$ on a grid of $\Lumtot$ values and
checked if the probability integrates to one using Simpson's
rule. With the exact expression, we obtain 0.997 ($n=10$) and 0.996
($n=80$), and 0.698 and 0.953 for the asymptotic expression. For
$n=10$, the large discrepancy arises mostly because the positivity
constraint $P(Q \leq 0 \cond n, \bml) = 0$ is not respected in
\refeq{ral} since the Gaussian approximation \refeq{rae} allows
negative $\lambda$ and $\Lumtot$, whereas the results \refeq{spk}
based on the $\GammaDist$ model imply $\lambda \geq 0$ and $Q \geq 0$.

For that reason, we discourage the asymptotic approximation for small
$n$ but we note that in general, the lack of numerical normalization
may not be a concern. For example, if \refeq{ral} is used as the
likelihood for a single bin in Bayesian parameter inference, the
normalization or evidence is
irrelevant. Following~\cite{tierney_accurate_1986}, we renormalized
the results to integrate to one for ease of comparison in
\reffig{asymptotic}.  We verified that for $n>80$, the two methods
agree even better than shown there and that the normalization of the
asymptotic result approaches one.

The main advantage of the approximation is computational efficiency. A
single evaluation of $P(Q \cond n, \bml)$ takes only about 100 calls
to the integrand of \refeq{ral} regardless of the value of $n$. In
contrast, a the numerical integration of a single term for fixed $N$
in \refeq{spk} requires 1000s of function evaluations, and the number
of terms grows like $\sqrt{n}$ for a given accuracy of
$P(Q \cond n, \bml)$.

%%%%%%%%%%%%%%%%%%%%%%%%%%%
\hans{Some comments on the numerical values of hyperparameters $a$ in
  Poisson, $a,b$ in inverse gamma needed.} %
\hans{I would omit the text and calculations below. Apart from the
  extra parameters, the main reason is that the Gaussian prior in
  $\mu$ is not consistent with the fact that all luminosities are
  positive, while the gaussian permits negative $\mu$. This may not be
  important numerically, but the uniform prior in $\mu$ does
  circumvent this problem and is moreover less informative than the
  Gaussian prior. } %
\hans{Note also that letting $\mu_0{=}0$ and $ n_0\to 0$ below implies
  a prior gaussian of infinite width centered on zero, i.e.\ half of
  the values of $\mu$ will be negative. Could this be the reason why
  the numerics pick up a factor two?}%
\fred{No, what I used before was equivalent to a uniform prior that
  extends into the range $\mu<0$. The normalization cancels out and
  the negative region doesn't contribute. With the uniform prior, you
  assume that your upper cut-off on $\mu$ is large enough to contain
  the posterior but at the same time large values don't actually
  contribute so it doesn't matter if you include them or not. That
  means I can't pick up a factor of two there, it would only change
  the normalization but not the \emph{shape} of $p(Q)$} \fred{The
  error is fixed and I remove the next sections as Hans suggested.}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\subsection{Moments of the luminosity prediction}\label{sec:predict-moments}

\hans{H: If threefold integrals must be carried out for the
  expectation values, why not do so for Eq (ral) directly rather than
  just the expectation values?} %

\hans{H: I would agree to remove this section; I do not see value added.}

@@

\fred{Our implementation is available online on \href{https://github.com/tardis-sn/XXX}{github}}

>>>>

@@

We return to the general formulation of \refeq{sll} and re-arrange its
factors into
\begin{align}
  \lleq{asc}
  p(Q\cond n,\bml)
  &= \int \rmdx{\lambda}\; \rmdx{\bmphi}\;
    p(\lambda\cond n)\;
    p(\bmphi\cond n,\bml)\;
    p(Q\cond \lambda,\bmphi), \\
  \lleq{asd}
  p(Q\cond \lambda,\bmphi)
  &= \sum_N
    p(N\cond \lambda)\;
    \int \rmdx{\bmL}\;
    p(Q\cond N,\bmL)\;
    p(\bmL\cond N,\bmphi),
\end{align}
with $p(\bmL\cond N,\bmphi) = \prod_{j=1}^N p(L_j\cond\bmphi)$ as
before. Again all quantities are related to a single bin in frequency.
Writing the moments and variance of $L$ as\footnote{The $k$-th moment
  of variable $x$ is defined as the expectation value $\mu_k =
  \expect{x^k} = \int \rmdx{x}\,x^k\,p(x),\; k = 0,1,2,\ldots$.} %
\begin{align}
  \lleq{ram}
  \mu =  \mu_1 &\equiv \expect{\Lum}, \qquad
  \mu_2 \equiv \expect{\Lum^2}, \\
  \sigma^2 &\equiv \variance{\Lum}
  = \expect{\Lum^2} - \expect{\Lum}^2,
\end{align}
it is easy to show that, for the Poissonian compound process
(\ref{asd}), the first moment and variance of $Q$ are related to those
of $L$ by
\begin{align}
  \lleq{raa}
  \expect{Q} = \lambda \mu, \qquad
  \variance{Q} = \lambda \mu_2 = \lambda (\mu^2 + \sigma^2) \,.
\end{align}
for any probability $p(L\cond\bmphi)$.
%
Higher-order quantities such as the skewness and kurtosis\footnote{
  The variance is the second so-called \textit{cumulant}. The relation
  for all orders is that the $k$-th order cumulant $\kappa_q(Q)$ is
  equal to $\lambda$ times the $k$-th order moment of $L$,
  $\mu_k$. The skewness of $p(Q\cond \lambda,\bmphi)$ is therefore
  $\kappa_3(Q)/\kappa_2(Q)^{3/2} = \lambda \mu_3/(\lambda \mu_2)^{3/2}
  = \lambda^{-1/2} \mu_3/\mu_2^{3/2}$.}  %
are suppressed by powers of $\lambda^{-1/2}$.
%%%%%%%%%%%
\hans{H comment (not for text): The magnitude of the correction to
  Eq.~(\ref{rab}) due to $\kappa_3$ and higher orders is unclear.}
%%%%%%%%%%%
From Theorem 4.3.1 of Ref.~\cite{bening2002generalized}, we conclude
that the asymptotic distribution of $\Lumtot$ when
$\lambda \to \infty$, given the first and second moment of $\Lum$, is
a Gaussian or normal distribution,
\begin{align}
  \lleq{rab}
  p(\Lumtot \cond \lambda,\bmphi)
  &= \GaussianDist \left( \Lumtot \cond \expect{Q}, \variance{Q} \right) %
    = \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\mu^2 {+} \sigma^2) \right).
\end{align}
This drastic simplification eliminates both the sum over $N$ and the
integral over $\bmL$ in \refeq{asd}.

It remains to find the integrals over $\lambda$ and $\bmphi$ in
\refeq{asc}. As before, the posterior for $\lambda$ is the Gamma
distribution of \refeq{pms}. %
%
To determine $p(\bmphi\cond n,\bml)$, we note that the result
(\ref{rab}) also simplifies the calculation of the parameter posterior
$p(\bmphi\cond n,\bml)$ since, independently of the actual functional
model for $p(L\cond\bmphi)$ the only relevant information is contained
in the first moment $\mu$ and variance $\sigma^2$.  As set out in
Jaynes~\cite[Ch. 7.10]{jaynes2003probability}, knowledge of $\mu$ and
$\sigma^2$ implies that we can assign a Gaussian distribution to
$p(L\cond\bmphi)$
\begin{align}
  \lleq{rae}
  p(\Lum \cond \bmphi) =
  p(\Lum \cond \mu, \sigma^2 ) &= \GaussianDist(\Lum \cond \mu, \sigma^2)
\end{align}
even though the actual data histograms may not be Gaussian. The reason
is that a Gaussian represents our state of knowledge \emph{before}
data $\bmy$ is observed. With this assignment,
$\bmphi =(\mu,\sigma^2)$ and the posterior becomes, using Bayes'
Theorem,
\begin{align}
  \lleq{rac}
  p(\mu,\sigma^2\cond n,\bml)
  &= \frac{
    \GaussianDist(\bml\cond\mu,\sigma^2)\;
    p(\mu\cond n)\;
    p(\sigma^2\cond n)}
  {p(\bml\cond n)}
\end{align}
In keeping with prior information that luminosities are always
positive, we assign an improper uniform prior
$p(\mu\cond\mu_{\rm max}) = 1/\mu_{\rm max}$ independent of $n$; as
always, the posterior is independent of
$\mu_{\rm max}$.
For $\sigma^2$, we use an $n$-independent Inverse Gamma
distribution prior with hyperparameters $(a_0,b_0)$ which reads
\begin{align}
  \lleq{rad}
  p(\sigma^2\cond a_0,b_0)
  = \InvGammaDist(\sigma^2\cond a_0,b_0)
  &= \frac{b_0^{a_0}}{\Gamma(a_0)} (\sigma^2)^{-a_0-1} \exp(-b_0/\sigma^2)
\end{align}
and is ``conjugate'' to the normal distribution likelihood of
\refeq{rae} in that the posterior for $(\mu,\sigma^2)$ is also a
Gaussian in $\mu$ and an Inverse Gamma in $\sigma^2$, but with
parameters updated by the data,
\begin{align}
  \lleq{rbd}
  p(\mu,\sigma^2\cond n,\bml,a_0,b_0)
  &= \GaussianDist(\mu\cond \mu_n,\sigma_n^2)
  \cdot
  \InvGammaDist(\sigma^2\cond a_n,b_n) \\
  \mu_n &= \langle\ell\rangle\\
  \sigma_n^2 &= \sigma^2/n \\
  a_n &= a_0 + \tfrac{1}{2}n \\
  b_n &= b_0 + \tfrac{1}{2} n\left( \langle\ell^2\rangle - \langle\ell\rangle^2\right)
\end{align}
with data sample moments $\langle\lum^k\rangle$ defined as
$(1/n) \sum_{j=1}^n \lum_j^k$. Since we assume $n \gg 1$, we may
choose $a_0 = b_0 = 0$ resulting in a noninformative prior on $\sigma^2$.
%
Inserting Eqs.~(\ref{pms}) and (\ref{rbd}) into \refeq{asc}, we obtain
\begin{align}
  \lleq{ral}
  p(\Lumtot \cond n,\bml)
  &= \int \rmdx{\lambda}\; \rmdx{\mu}\; \rmdx{\sigma^2}\;
  \GaussianDist \left( \Lumtot \cond \lambda \mu, \lambda (\sigma^2{+}\mu^2) \right)
  \times\\
  &\quad\times
  \GammaDist(\lambda \cond n{-}a{+}1 , 1) \cdot
  \GaussianDist(\mu\cond \mu_n,\sigma_n^2) \cdot
  \InvGammaDist(\sigma^2\cond a_n,b_n) \;
  \nonumber
\end{align}

@@

In \reffig{asymptotic}, we compare the asymptotic expression
\refeq{ral} with the ``exact'' expression \refeq{spk} where the
integrals over $\bmphi$ are done numerically to a requested relative
accuracy of $10^{-5}$ through the Cubature
package\cite{cubature,genz1980remarks,berntsen1991adaptive}.

For large $n$ where the asymptotic expression \refeq{ral} is valid,
asymptotic normality justifies the use of the Laplace or saddle-point
approximation~\cite[Ch. 27]{mackay2003information}. We implemented
\fredmargin{\href{https://github.com/tardis-sn/XXX}{github}}
the
above formulae in the julia language~\cite{julia14} and used Newton's
method from the \texttt{Optim.jl} package in combination with
automatic differentiation~\cite{RevelsLubinPapamarkou2016} to optimize
the respective integrand and to compute the Hessian at the mode as
required for the Laplace method. As initial values for optimization,
we slightly offset the sample estimates
\begin{align}
  \lleq{rat}
  \qty(\lambda, \mu, \sigma^2) = 1.0005 (n, \langle\lum\rangle, \langle\lum^2\rangle{-}\langle\lum\rangle^2)
\end{align}
to avoid issues if the optimizer cannot improve on the initial guess
and found convergence within $\order{10}$ steps.

As a test case, we generated
$\ell_i \sim \GammaDist(\alpha=1.5, \beta=60), i=1 \dots n$, and
computed $P(\Lumtot \cond n, \bml)$ on a grid of $\Lumtot$ values and
checked if the probability integrates to one using Simpson's
rule. With the exact expression, we obtain 0.997 ($n=10$) and 0.996
($n=80$), and 0.698 and 0.953 for the asymptotic expression. For
$n=10$, the large discrepancy arises mostly because the positivity
constraint $P(Q \leq 0 \cond n, \bml) = 0$ is not respected in
\refeq{ral} since the Gaussian approximation \refeq{rae} allows
negative $\lambda$ and $\Lumtot$, whereas the results \refeq{spk}
based on the $\GammaDist$ model imply $\lambda \geq 0$ and $Q \geq 0$.
%
\fred{Update values after Jeffreys implementing prior}

For that reason, we discourage the asymptotic approximation for small
$n$ but we note that in general, the lack of numerical normalization
may not be a concern. For example, if \refeq{ral} is used as the
likelihood for a single bin in Bayesian parameter inference, the
normalization or evidence is
irrelevant. Following~\cite{tierney_accurate_1986}, we renormalized
the results to integrate to one for ease of comparison in
\reffig{asymptotic}.  We verified that for $n>80$, the two methods
agree even better than shown there and that the normalization of the
asymptotic result approaches one. Integrating the asymptotic
expression with cubature instead of Laplace gave very similar results
but turns out to be less reliable and requires many more iterations.

The main advantage of the approximation is computational efficiency. A
single evaluation of $P(Q \cond n, \bml)$ with the Laplace methods
requires only about 100 calls to the integrand of \refeq{ral}
regardless of the value of $n$. In contrast, the numerical integration
of a single term for fixed $N$ in \refeq{spk} requires 1000s of
function evaluations, and the number of terms grows like $\sqrt{n}$
for a given accuracy of $P(Q \cond n, \bml)$.

@@

\subsection{Moments of the luminosity prediction}\label{sec:predict-moments}
\hans{H: I would agree to remove this section; I do not see value added.}

@@
